The reproductive and parental care transcriptome of the rock dove

### 2020-09-09 manuscript/elife_v1.Rmd
rm old files
vsd working, degs not working

Parental care is essential to many organisms’ reproductive strategy and predominates avian and mammalian groups (Clutton-Brock, 1991; Royle et al., 2012). Depending on the species, reproductive and parental behaviors may include territory establishment, courtship and mating, nest-building, nest defense, incubation, and the brooding and provisioning of young. The goal of these behaviors, either directly or indirectly, is the successful production of offspring resulting in a direct fitness benefit  (Reviewed in Santos, E.S.A, 2012). Underlying these behaviors are the anatomical and physiological changes that facilitate the transition into and across reproductive and parenting stages. These changes have fascinated biologists and inspired a prodigious body of research. Much of this research has focused on understanding reproduction and parental care from an endocrine and neuroendocrine perspective (reviewed in Ondrasek 2016, Kelly 2020, Wingfield, Silver, Buntin).
Reproduction and parental care is largely regulated by a set of three tissues: The Hypothalamus in the brain, the Pituitary, and the Gonads (testes and ovaries). Together, these tissues are often referred to as the ‘reproductive’ axis or the HPG. To activate reproduction, the HPG initiates an endocrine cascade starting with the production and release of gonadotropin-releasing hormone (GnRH) in the hypothalamus which, in turn,  stimulates the anterior pituitary to produce luteinizing hormone (LH) and follicle-stimulating hormone (FSH). LH and FSH travel to the gonads via the bloodstream where they induce gonadal growth and steroidogenesis of estradiol, testosterone and progesterone). These sex steroids have well established roles in mating behavior, nest-building, territory defense, and egg-laying in birds (reviewed in Buntin). Another key hormone, prolactin, is produced in the pituitary and is associated with the induction and maintenance of behaviors such as  incubation in birds and lactation (in doves and mammals) (birds: reviewed in Angelier, Austin et al. in prep, mammals: Austin and Word). Other hormones like vasopressin and oxytocin are related to affiliative behaviors (Yoshihara C, 2018) while dopamine, serotonin aggressive behaviors or avoidance (Rilling, JK, 2013; Adams DB, 2006). While these and many other hormones have been reported to play integral roles in various reproductive and parental care behaviors, we have yet to come close to conceptualizing the larger physiological and genomic coordination of events that yield such behaviors.

@@

```{r table1, echo =FALSE, message = F }
table1 <- read_csv("../results/table1.csv")
table1[is.na(table1)] <- " "
kable(table1,
      col.names = c('gene', 'bldg-lay',  'lay-inc3', 'inc3-inc9', 'inc9-inc17', 
                    'inc17-hatch', 'hatch-n5', 'n5-n9'),
      caption = "Changes in candidate gene expression. Significant chagns in gene expression are shown for X candidate genes. Positive and negative changes are denoted with (+) and (-) respectively for female (F) and male (M) hypothalamus (H), pituitary (P) and gonads (G). Comparisons are denoted in the column names. The first group is the reference group.") 
```

>>>>

Parental care is essential to many organisms’ reproductive strategy and predominates avian and mammalian groups (Clutton-Brock, 1991; Royle et al., 2012). Depending on the species, reproductive and parental behaviors may include territory establishment, courtship and mating, nest-building, nest defense, incubation, and the brooding and provisioning of young. The goal of these behaviors, either directly or indirectly, is the successful production of offspring resulting in a direct fitness benefit  (Reviewed in Santos, E.S.A, 2012). Underlying these behaviors are the anatomical and physiological changes that facilitate the transition into and across reproductive and parenting stages. These changes have fascinated biologists and inspired a prodigious body of research. Much of this research has focused on understanding reproduction and parental care from an endocrine and neuroendocrine perspective (reviewed in Ondrasek 2016, Kelly 2020, Wingfield, Silver, Buntin).

Reproduction and parental care is largely regulated by a set of three tissues: The Hypothalamus in the brain, the Pituitary, and the Gonads (testes and ovaries). Together, these tissues are often referred to as the ‘reproductive’ axis or the HPG. To activate reproduction, the HPG initiates an endocrine cascade starting with the production and release of gonadotropin-releasing hormone (GnRH) in the hypothalamus which, in turn,  stimulates the anterior pituitary to produce luteinizing hormone (LH) and follicle-stimulating hormone (FSH). LH and FSH travel to the gonads via the bloodstream where they induce gonadal growth and steroidogenesis of estradiol, testosterone and progesterone). These sex steroids have well established roles in mating behavior, nest-building, territory defense, and egg-laying in birds (reviewed in Buntin). Another key hormone, prolactin, is produced in the pituitary and is associated with the induction and maintenance of behaviors such as  incubation in birds and lactation (in doves and mammals) (birds: reviewed in Angelier, Austin et al. in prep, mammals: Austin and Word). Other hormones like vasopressin and oxytocin are related to affiliative behaviors (Yoshihara C, 2018) while dopamine, serotonin aggressive behaviors or avoidance (Rilling, JK, 2013; Adams DB, 2006). While these and many other hormones have been reported to play integral roles in various reproductive and parental care behaviors, we have yet to come close to conceptualizing the larger physiological and genomic coordination of events that yield such behaviors.

@@

```{r table1, echo =FALSE, message = F }
table1 <- read_csv("../results/table1.csv")
table1[is.na(table1)] <- " "
kable(table1,
      col.names = c('gene', 'bldg-lay',  'lay-inc3', 'inc3-inc9', 'inc9-inc17', 
                    'inc17-hatch', 'hatch-n5', 'n5-n9'),
      caption = "Changes in candidate gene expression over the course of parental care in rock doves. Significant changes in gene expression are shown for X candidate genes. Positive and negative changes are denoted with (+) and (-) respectively for female (F) and male (M) hypothalamus (H), pituitary (P) and gonads (G). Comparisons are denoted in the column names. The first group is the reference group.") 
```

```{r table2, echo =FALSE, message = F }
table2 <- read_csv("../results/table2.csv")
table2[is.na(table2)] <- " "
kable(table2,
      caption = "Changes in candidate gene expression in response to removal. Significant changess in gene expression are shown for X candidate genes. Positive and negative changes are denoted with (+) and (-) respectively for female (F) and male (M) hypothalamus (H), pituitary (P) and gonads (G). Comparisons are denoted in the column names. The first group is the reference group.") 
```

```{r table3, echo =FALSE, message = F }
table3 <- read_csv("../results/table3.csv")
table3[is.na(table3)] <- " "
kable(table3,
      caption = "Changes in candidate gene expression in response to replacement. Significant changess in gene expression are shown for X candidate genes. Positive and negative changes are denoted with (+) and (-) respectively for female (F) and male (M) hypothalamus (H), pituitary (P) and gonads (G). Comparisons are denoted in the column names. The first group is the reference group.") 
```

levelsmanip
